---
title: "IncompleteCounterbalancing"
author: "Ihor Markevych"
date: "10/25/2019"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task

You want to conduct an experiment to test the effect of a new drug for treating children with attention deficit
disorder. You have decided to test six different amounts of the new drug-5, 7, 9, 11, 13, and 15 mg. The
parents of 40 children with attention deficit disorder have volunteered their child to participate in the study.  
  
You know that using this procedure might result in either carryover or order effects, so you want to
counterbalance the order of presentation of the six dosages. Construct the different counterbalanced order of
treatment conditions using the incomplete counterbalancing technique in r markdown

# Functions

## Incomplete Counterbalancing

```{r}
mod <- function(n, m){
   n - m * floor(n / m)
}

generateFirstSequence <- function(n) {
    seq  <- c(1)
    i <- 1
    while(i < n - i + 1){
        i <- i + 1
        seq <- c(seq, i)  
        if(i == n - i + 2){
            break
        }
        seq <- c(seq, n - i + 2) 
    }
    return(seq)
}

incompleteCounterBalance <- function(n){
    res <- list(generateFirstSequence(n))
    for(i in 2:n){
        res <- append(res, list(mod(res[[length(res)]], n) + 1))
    }
    
    if(mod(n,2) == 1){
        for(i in 1:n){
            res <- append(res, list(rev(res[[i]])))
        }
    }
    
    return(res)
}
```

## Random Assignment

```{r}
randomAssignment <- function(n_groups, n_participants, returnFlat = FALSE){
    treatments <- sample(n_groups, n_participants, replace = TRUE)
    groups <- list()
    for(i in unique(treatments)){
        groups <- append(groups, list(which(treatments == i)))
    }
    
    if(returnFlat){
        return(treatments)
    }
    
    return(groups)
    
}
```

## Printing Treatments 

```{r}
printTreatments <- function(treatments, order){
    i <- 1
    for(seq in order){
        cat('Sequence', i, '\t|\t')
        for(tr in seq){
            cat(treatments[tr],'\t')
        }
        cat('\n')
        i <- i + 1
    }
}
```

## Print Groups

```{r}
printGroups <- function(groups){
    i <- 1
    for(group in groups){
        cat('Participants of sequence', i, '\n')
        cat(group, '\n\n')
        i <- i + 1
    }
}
```

# Results

## Sequences

```{r}
treatments <- c('5mg', '7mg', '9mg', '11mg', '13mg', '15mg')

printTreatments(treatments, incompleteCounterBalance(length(treatments)))
```

## Groups

Each child-participant should be assigned a number. Next table represents to which treatment sequence each participant should be allocated.

```{r}
printGroups(randomAssignment(length(treatments), 40))
```